Verified Nanopasses for Compiling Conditionals

### 2025-08-12 paper.tex
spell check

Towards developing a course on verified compilation, I am interested
in proving correctness of the compiler presented in the
\emph{Essentials of Compilation}
textbook~\citep{Siek:2023tr,Siek:2023ue}. While it does not include
proofs of correctness for its compiler passes, the textbook is
organized in a way that might facilitate the proofs of correctness.
%
\emph{Essentials of Compilation} takes an \emph{incremental} approach
to teaching compilers, that is, it guides the students through the
creation of multiple compilers for a sequence of languages, each of
which adds a language feature to the previous language. For the
language with Boolean expressions and conditionals, the Explicate
Control pass of the compiler uses the one-function per context
approach of Danvy to translate into an x86-flavored register transfer
language.
%
In addition to being incremental, \emph{Essentials of Compilation}
employs the \emph{nanopass} approach, breaking up the compiler into
many small passes that are easier for students to
digest~\citep{Sarkar:2004fk}.
%
There are many other pedogogical treatments of
compilation~\citep{Appel:2003fk,Sumii:2005aa,Aho:2006wb,Cooper:2011aa,Grune:2012aa}
but none of them take the incremental approach.

@@

% LocalWords:  sigplan tex docstrip dtx bibtex TeXCount TC envir Siek
% LocalWords:  displaymath Nanopasses authornote authornotemark Tobin
% LocalWords:  rv Trovato al Agda nanopasses Olivier Danvy's ccs desc
% LocalWords:  nanopass Ichiro Aho qf gotos datatypes PEYTONJONES fk
% LocalWords:  Danvy arity Dybvig's Dybvig aa Chez Olivier's CompCert
% LocalWords:  fe Rocq RTLgen CminorSel CompCert's CakeML stackLang tr
% LocalWords:  labLang Kumar ue Sarkar whac de Bruijn kx Eq Exp Num
% LocalWords:  Bool Var Int Env interp exp LIf immediates Atm atm mon
% LocalWords:  LMonIf IR AST Prog StateImp cmpq je jmp CExp CStmt CFG
% LocalWords:  IfEq CProg prog Arg Reg Dest Inst MovQ SubQ CmpQ JmpEq
% LocalWords:  ReadInt rax StateX EFLAGS var inst rco cmplx suc IL ie
% LocalWords:  retB thenB pred GHC thn eval Agda's transl condexpr
% LocalWords:  stmt arg dest regs DestOK trans

>>>>

Towards developing a course on verified compilation, I am interested
in proving correctness of the compiler presented in the
\emph{Essentials of Compilation}
textbook~\citep{Siek:2023tr,Siek:2023ue}. While it does not include
proofs of correctness for its compiler passes, the textbook is
organized in a way that might facilitate the proofs of correctness.
%
\emph{Essentials of Compilation} takes an \emph{incremental} approach
to teaching compilers, that is, it guides the students through the
creation of multiple compilers for a sequence of languages, each of
which adds a language feature to the previous language. For the
language with Boolean expressions and conditionals, the Explicate
Control pass of the compiler uses the one-function per context
approach of Danvy to translate into an x86-flavored register transfer
language.
%
In addition to being incremental, \emph{Essentials of Compilation}
employs the \emph{nanopass} approach, breaking up the compiler into
many small passes that are easier for students to
digest~\citep{Sarkar:2004fk}.
%
There are many other pedagogical treatments of
compilation~\citep{Appel:2003fk,Sumii:2005aa,Aho:2006wb,Cooper:2011aa,Grune:2012aa}
but none of them take the incremental approach.

@@

% LocalWords:  sigplan tex docstrip dtx bibtex TeXCount TC envir Siek
% LocalWords:  displaymath Nanopasses authornote authornotemark Tobin
% LocalWords:  rv Trovato al Agda nanopasses Olivier Danvy's ccs desc
% LocalWords:  nanopass Ichiro Aho qf gotos datatypes PEYTONJONES fk
% LocalWords:  Danvy arity Dybvig's Dybvig aa Chez Olivier's CompCert
% LocalWords:  fe Rocq RTLgen CminorSel CompCert's CakeML stackLang tr
% LocalWords:  labLang Kumar ue Sarkar whac de Bruijn kx Eq Exp Num
% LocalWords:  Bool Var Int Env interp exp LIf immediates Atm atm mon
% LocalWords:  LMonIf IR AST Prog StateImp cmpq je jmp CExp CStmt CFG
% LocalWords:  IfEq CProg prog Arg Reg Dest Inst MovQ SubQ CmpQ JmpEq
% LocalWords:  ReadInt rax StateX EFLAGS var inst rco cmplx suc IL ie
% LocalWords:  retB thenB pred GHC thn eval Agda's transl condexpr wb
% LocalWords:  stmt arg dest regs DestOK trans Lawall Appel Sumii cg
% LocalWords:  Grune Isabelle Hatcliff vn Kelsey uq Birkedal movq int
% LocalWords:  basicstyle mathescape callq subq
